The pragmatics of perspective-taking

### 2018-05-14 writing/journal_manuscript/manuscript.tex
Add expt. 2 results

Finally, we consider the link between our speaker and listener results. If listeners have a well-tuned pragmatic expectation about the speaker's likely level of informativity, they may rationally neglect the speaker's perspective and simply pick the object that fits the description best. To what extent does the informativity of speaker utterances account for heterogeneity in listener errors across items in the two conditions? As predicted by a perspective-negligence account, we found a strong negative correlation between informativity and error rates: listeners make fewer errors when utterances are a good fit for the target relative to the distractor ($\rho = -0.8$, bootstrapped 95\% CI $= [-0.89, -0.69]$; Fig. \ref{fig:exp1speaker}B). %  We used a mixed effects logistic regression model to estimate the effect of target and distractor fit on the probability of making a critical error in both experiments, including a random intercept for game ID. We found that participants are less likely to make errors when informativity is higher. Furthermore, a model including target fit and distractor fit in addition to item-level fixed effects is significantly better than a model including item alone, $\chi^2(2) = XXX, p < YYY$, implying that speaker informativity captures residual variance beyond the item-wise effects reported above.
\todo[inline]{rxdh: I think this bootstrapped correlation stat at the objectSet-condition level makes this point well enough for our purposes (it's a pretty huge effect...), but I could also do a fancier thing boostrapping logistic coefficients at the trial-by-trial level if necessary. }
\todo[inline]{rxdh: also, is it worth having a short appendix with more details on the multi-stage bootstrapping procedure? it's very straightforward but takes a few sentences to explain\dots}

@@

At the same time, however, our \emph{speaker} results suggest a deeper explanation for this phenomenon of perspective-neglect. On filler trials in the scripted condition and all trials in the unscripted condition, we noted a tendency toward additional, possibly unnecessary, precision in descriptions. For example, instead of the scripted ``move the brush one square to the right'' when there is only one brush in their display, speakers instead gave descriptions like ``move the hairbrush with two lines on the handle to the right.'' 
Perhaps directors were taking the time to make more precise descriptions because they believed it was contextually relevant: both parties know that there are hidden objects in the environment, thus increasing the chance of miscommunication from imprecise descriptions.

  \begin{figure}[t!]
 \begin{center}
 \vspace{-.5cm}
 \includegraphics[scale=.25]{figures/Exp2_design}
 \caption{Within-pair factorial design used in Exp.~2. In the common condition, we reproduce a simple pragmatic reference contrast: when more than one star is present in common ground, speakers are expected to be more informative. On hidden versions of these condition, the same set of objects is visible, but some additional squares are occluded. If pragmatic directors represent uncertainty over what the matcher sees behind these squares, we predict that they will produce more informative utterances.}
 \vspace{-.5cm}
 \label{fig:exp2design}
 \end{center}
 \end{figure}

@@

In Experiment 2, we more directly test the hypothesis that speakers naturally produce more informative referring expressions in more uncertain contexts where items are occluded. For this experiment, we move to a more systematic stimulus domain. The idiosyncracies we observed among the linguistic items used in Expt. 1 make it difficult to normatively assess pragmatic increases in informativity for a given target. For example, when unscripted participants used an additional modifier (`cassette tape') to describe the target tape in one item, it happened to be highly informative relative to the distractor (a roll of tape). Adding a similar modifier (`hair brush') to describe the target in another item, however, was ineffective because both the target and distractor are hair brushes that are confusable at the subordinate level (one is round and one is flat). Unsurprisingly, then, for

@@

We recruited 200 participants (100 pairs) from Amazon Mechanical Turk. All participants were native English speakers.

@@

Stimuli were sampled from a more systematic domain than the idiosyncratic objects used in Exp.~1. They varied along three discrete features: \emph{shape}, \emph{texture}, and \emph{color}. Each feature took four discrete values, giving a space of 64 possible objects. We placed these objects in a smaller $3 \times 3$ grid, and used a pure \emph{reference game} setup where listeners had to click the intended referent as opposed to following a movement instruction.

Our design is entirely within-pair. Each trial is sampled from a factorial 2 x 2 design manipulating (1) the presence or absence of occlusions (`hidden' vs. `no hidden'), and (2) the closeness of visible distractors to the target (`close' vs. `far'; see Fig. \ref{fig:exp2design}). On 'hidden' trials, two cells are covered in curtains such that only the matcher can see if there is an object in the cell; on 'no hidden' trials, all cells are in common ground. On 'far' trials, the target is the only object with a particular shape; on 'close' trials, there is also a visible distractor sharing the target's shape, differing only in color or texture. Each trial type appeared 6 times for a total of 24 trials, and the sequence of trials was pseudo-randomized such that no trial type appeared more than twice in each

 To test how speakers flexibly modify their informativity in response to the uncertainty posed by

\subsection{Results}
 \begin{figure}[t!]
 \begin{center}
 \vspace{-.25cm}
 \includegraphics[scale=1.25]{figures/Exp2_results.pdf}
 \vspace{-.5cm}
 \caption{Results for Exp.~2. (A) Speakers used significantly more words when occlusions were present, and effect is even greater than the standard pragmatic effect of being in a close vs. far context (B) Utterances broken out by feature mentioned. All error bars are bootstrapped 95\% confidence intervals.}
 \vspace{-.5cm}
 \label{fig:exp2results}
 \end{center}
 \end{figure}

Our primary is the length (in words) of their naturally produced referring expressions in different contexts. \ref{fig:exp2results}A

As a secondary measure (intended for use in a cognitive model), we will also annotate each utterance by the number of object features mentioned (e.g. shape, texture, color). \ref{fig:exp2results}B

>>>>

Finally, we consider the link between the speaker and listener behavior we observed across condition. To what extent does the informativity of speaker utterances account for the substantial variance across items we observed in listener errors in the two conditions? While a naive perspective-taking account would predict no relationship, a perspective-negligence account predicts that listeners simply pick the object that fits the description best regardless of whether it is hidden or not. Indeed, we found a strong negative correlation between informativity and error rates: listeners make fewer errors when utterances are a better fit for the target relative to the distractor ($\rho = -0.8$, bootstrapped 95\% CI $= [-0.89, -0.69]$; Fig. \ref{fig:exp1speaker}B). %  We used a mixed effects logistic regression model to estimate the effect of target and distractor fit on the probability of making a critical error in both experiments, including a random intercept for game ID. We found that participants are less likely to make errors when informativity is higher. Furthermore, a model including target fit and distractor fit in addition to item-level fixed effects is significantly better than a model including item alone, $\chi^2(2) = XXX, p < YYY$, implying that speaker informativity captures residual variance beyond the item-wise effects reported above.
\todo[inline]{rxdh: I think this bootstrapped correlation stat at the objectSet-condition level makes this point well enough for our purposes (it's a pretty huge effect...), but I could also do a fancier thing boostrapping logistic coefficients at the trial-by-trial level if necessary. }
\todo[inline]{rxdh: also, is it worth having a short appendix with more details on the multi-stage bootstrapping procedure? it's very straightforward but takes a few sentences to explain\dots}

@@

At the same time, however, our \emph{speaker} results suggest a deeper explanation for the phenomenon of perspective-neglect. We found a tendency toward additional, possibly unnecessary, informativity in speaker descriptions. For example, instead of the scripted ``move the \emph{brush} one square to the right'' which would be sufficient with only one brush in their display, speakers instead chose to give descriptions like ``move the \emph{hairbrush with two lines on the handle} to the right.'' 
Perhaps speakers were taking the time to make more precise descriptions than required by their own immediate display because they believed it was contextually relevant. 
In particular, speakers know that there additional objects of unknown identity in their partner's field of view, potentially increasing the risk of miscommunication from imprecise descriptions. 
If natural speakers do adaptively produce more informative descriptions in context, then pragmatic listeners expecting speakers to do so may be justified in weighting considerations of the speaker's visual perspective less strongly.

@@

In Experiment 2, we more directly test the hypothesis that speakers adaptively increase the informativity of their referring expressions in more \emph{uncertain} contexts where items are occluded. The idiosyncracies we observed among the linguistic items used in Expt. 1 make it difficult to normatively assess pragmatic `increases' in informativity for a given target. For example, when unscripted participants used an additional modifier (`cassette tape') to describe the target tape in one item, it happened to be highly informative relative to the distractor (a roll of tape). Adding a similar modifier (`hair brush') to describe the target in another item, however, was ineffective because both the target and distractor are hair brushes that are confusable at the subordinate level (one is round and one is flat). For this experiment, then, we move to a more systematic, feature-based stimulus domain and manipulate the presence or absence of occluders.

@@

We recruited 204 participants (102 pairs) from Amazon Mechanical Turk. After we removed 7 games that disconnected part-way through and 12 additional games according to our pre-registered exclusion criteria (due to being non-native English speakers, reporting confusion about the instructions, or clearly violating the instructions), we were left with a sample of 84 full games.

@@

  \begin{figure}[t!]
 \begin{center}
 \vspace{-.5cm}
 \includegraphics[scale=.35]{figures/Exp2_design.pdf}
 \caption{Within-pair factorial design used in Exp.~2. In the common condition, we reproduce a simple pragmatic reference contrast: when more than one star is present in common ground, speakers are expected to be more informative. On hidden versions of these condition, the same set of objects is visible, but some additional squares are occluded. If pragmatic directors represent uncertainty over what the matcher sees behind these squares, we predict that they will produce more informative utterances.}
 \vspace{-.5cm}
 \label{fig:exp2design}
 \end{center}
 \end{figure}

Stimuli varied along three discrete features: \emph{shape}, \emph{texture}, and \emph{color}. Each feature took four discrete values, giving a space of 64 possible objects. We placed these objects in a smaller $3 \times 3$ grid, and used a pure \emph{reference game} setup where listeners had to click the intended referent as opposed to following a movement instruction.

Our design is entirely within-pair. Each trial is sampled from a factorial 2 x 2 design manipulating (1) the presence or absence of occlusions (`hidden' vs. `no hidden'), and (2) the closeness of \emph{visible} distractors to the target (`close' vs. `far'; see Fig. \ref{fig:exp2design}). On `hidden' trials, two cells were covered in curtains such that only the matcher can see if there is an object in the cell; on 'no hidden' trials, all cells are in common ground. On 'far' trials, the target is the only object with a particular shape; on 'close' trials, there is also a visible distractor sharing the target's shape, differing only in color or texture. Each trial type appeared 6 times for a total of 24 trials, and the sequence of trials was pseudo-randomized such that no trial type appeared more than twice in each block of eight trials.

 \begin{figure}[t!]
 \begin{center}
 \vspace{-.25cm}
 \includegraphics[scale=1.25]{figures/Exp2_results.pdf}
 \vspace{-.5cm}
 \caption{Results for Exp.~2. (A) Speakers used significantly more words when occlusions were present, and effect is even greater than the standard pragmatic effect of being in a close vs. far context (B) Utterances broken out by feature mentioned. All error bars are bootstrapped 95\% confidence intervals.}
 \vspace{-.5cm}
 \label{fig:exp2results}
 \end{center}
 \end{figure}

Listener errors were rare and unstructured: 88\% of listeners made one or fewer errors, and there was no significant difference in error rates across the four conditions ($\chi^2(3) = 1.23, p = 0.74$). Because this experiment was designed to investigate speaker pragmatics, we focus our attention on how speaker behavior adjusted flexibly from context to context.

Our primary measure for speakers is the length (in words) of naturally produced referring expressions. We conducted a mixed-effect regression of context and occlusion on number of words with maximal random effect structure containing intercept, slopes, and interaction. First, we examined the \emph{simple} effect of close vs. far contexts in trials with no occlusions. We found that speakers used significantly more words on average when there was a distractor in context that shared the same shape as the target ($b = 0.56, t = 5.1, p < 0.001$; see Fig. \ref{fig:exp2results}A). This replicates the findings of classic studies with similar designs \cite[e.g] {BrennanClark96_ConceptualPactsConversation, GrafEtAl16_BasicLevel, MonroeEtAl17_ColorsInContext}. Next, we turn to the simple effect of occlusion in far contexts (which are most similar to the displays used in Expt.~1). The average speaker used 1.25 additional words when they knew their partner could potentially see additional objects ($t = 7.5, p < 0.001$). Finally, we found a significant interaction ($b = -0.49, t = 3.8, p <0.001$) where the effect of occlusion was larger in far contexts, likely indicating a ceiling on the level of informativity possible in our simple stimulus space.

What are these additional words spent on describing? As a secondary analysis, we annotated each utterance by which object features were mentioned (shape, texture, color). Speakers nearly always mentioned shape  (e.g. `star', `triangle') as the head noun of their referring expression regardless of context, so differences in words across conditions must be due to providing additional information about color and texture. We ran separate mixed-effect logistic regressions for color and texture predicting mention from context; due to well-known convergence issues in logistic models, the maximum random effect structure supported by our data contains only speaker-level intercepts and slopes for occlusion. In both cases, we found simple effects of occlusion in far contexts ($b = 1.33, z = 2.9, p = 0.004$ for color; $b = 2.1, z = 5.7, p < 0.001$ for texture, see Fig. \ref{fig:exp2results}B). In other words, in displays like the one in Fig. \ref{fig:exp2results}A where the target was the only star, speakers were slightly more likely to produce the star's color---and substantially more likely to produce its texture---when there were occlusions present even though the visible objects were exactly the same.

\todo[inline]{Say something about the small population of 10 or so speakers who adopted the strategy of mentioning every feature every time?}
